---
title: "Using Poistan"
author: "Ken Butler"
date: '2015-05-06'
output: html_document
---

Illustrating the use of `Poistan`
=================================

We illustrate how the package works using the two included data sets. 

The fictitious data
-------------------

The data set `m` is made-up, but illustrates the kind of data that the package uses as input:

```{r}
library(poistan)
data(m)
m
```

The data frame has four columns: the names of the home and away teams, and the number of goals scored by each in the game where they met. Thus, each row represents a game.

If necessary, the Stan model object must be compiled first:

```{r,cache=TRUE}
poisson.sc=poistan:::compiled_model()
poisson.sc
```

As long as the object `poisson.sc` is kept around, this only ever needs to be done once (eg.\ when the package is first installed).

Before fitting the model, we have to create a design matrix `X` and  response matrix `y` that represent the teams by numbers (and maintain a separate list of team names). This is done using `makexy`:

```{r}
xy=poistan:::makexy(m)
xy
```

